Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-02 paper.tex
Fixed #13

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years; as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of different
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration challenge, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration challenge, robustness,
extensibility, scalability and other specific features. Criteria break
down to specific metrics where \emph{lower value indicates better
  performance}. For each criterion, a global score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that they can be
summed. The criteria and metrics are explained hereafter.

@@

\cellcolor[HTML]{EEEEEE}\textbf{Integration challenge}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\

@@

\emph{Integration challenge} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain a
global score that measures the
total number of software pieces to develop, including interactions and
components. It may be used to measure the development effort required
to build the architecture.

@@

\paragraph{Integration challenge} This architecture does not require any
component in addition to the science gateway and infrastructure
(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{I$_2$=5}).

@@

\paragraph{Integration challenge} Service invocation requires a workflow service
in addition to the science gateway and infrastructure
(\texttt{I$_1$}=3). The architecture involves 6 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$},
\texttt{c$_2$} and \texttt{d} (\texttt{I$_2$}=6).

@@

\paragraph{Integration challenge} Task encapsulation requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Integration challenge} The pool model requires a workflow pool and an
agent in addition to the science gateway and infrastructure
(\texttt{I$_1$=4}). It involves 7 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$}, \texttt{c$_2$}, \texttt{d}
and \texttt{f} (\texttt{I$_2$}=7).

@@

\paragraph{Integration challenge} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.}

@@

\paragraph{Integration challenge} Workflow import with service invocation
requires the same components as for service invocation
(\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
for workflow import (\texttt{I$_1$}=7).

>>>>

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years; as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of different
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration complexity, robustness,
extensibility, scalability and other specific features. Criteria break
down to specific metrics where \emph{lower value indicates better
  performance}. For each criterion, a global score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that they can be
summed. The criteria and metrics are explained hereafter.

@@

\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\

@@

\emph{Integration complexity} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain a
global score that measures the
total number of software pieces to develop, including interactions and
components. It may be used to measure the development effort required
to build the architecture.

@@

\paragraph{Integration complexity} This architecture does not require any
component in addition to the science gateway and infrastructure
(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{I$_2$=5}).

@@

\paragraph{Integration complexity} Service invocation requires a workflow service
in addition to the science gateway and infrastructure
(\texttt{I$_1$}=3). The architecture involves 6 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$},
\texttt{c$_2$} and \texttt{d} (\texttt{I$_2$}=6).

@@

\paragraph{Integration complexity} Task encapsulation requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Integration complexity} The pool model requires a workflow pool and an
agent in addition to the science gateway and infrastructure
(\texttt{I$_1$=4}). It involves 7 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$}, \texttt{c$_2$}, \texttt{d}
and \texttt{f} (\texttt{I$_2$}=7).

@@

\paragraph{Integration complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.}

@@

\paragraph{Integration complexity} Workflow import with service invocation
requires the same components as for service invocation
(\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
for workflow import (\texttt{I$_1$}=7).
